---
title: "Intro to PUMSutils"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Intro to PUMSutils}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE)
library(dplyr)
library(PUMSutils)
```

## ACS PUMS Data

The American Community Survey (ACS) is an annual survey conducted by the U.S. Census Bureau, covering many of the same topics as the decennial census. It is a 1% sample, and it comes in 2 parts: a housing survey of household units and a population survey of people. ACS data is available in several forms. Online, you can get custom tables from [American FactFinder](https://factfinder.census.gov/faces/nav/jsf/pages/searchresults.xhtml?refresh=t). Microdata is available through the Public Use Microdata Sample, or PUMS. This contains actual responses to the ACS, one per row. This allows essentially any query to be formed, not just those provided by AFF. However, the data is harder to work with, since every query requires doing tabulation. The `PUMSutils` package reduces this part to function calls.    

The most important thing to know about the ACS PUMS data is that it is a *weighted* sample. 
Each row has an expansion weight indicating the number of households or people represented by this row.
In addition to the main expansion weight, there are 80 weight replicates, which are used in a bootstrap-like procedure for computing standard errors.

In the `PUMSutils` package, a partial 2016 ACS PUMS housing record for Washington State is included as `wa.house16` and Census Bureau-provided ground truth estimates for some Washington State housing characteristics are provided as `wa.gold16`. 
Below we show part of the `wa.gold16` data.

```{r gold, echo=FALSE}
# Ground truth for Washington State
wa.gold16[c(1, 3, 4, 9),]
```

Here there are 4 characteristics shown, total housing units, the number of owned
units, the number of rented units, and the number of vacant units.
The column `pums_est_16` has point estimates of these characteristics.
The third column, called `pums_se_16`, has standard errors.

If we were not using `PUMSutils`, we would compute a point estimate of the housing unit count like this:

```{r noPUMSutils}
# Sum the weights to estimate a count
sum(wa.house16$WGTP)
```

Proportions, various statistics and standard errors can also be computed, 
but the calculations are more involved. 
The `PUMSutils` package simplifies these calculations.

## PUMSutils

The `PUMSUtils` package has functions for computing estimates and standard errors for counts, proportions, and means, as well as computing tables and point estimates for medians and quantiles. It also contains utility functions for loading PUMS datasets, renaming the numerous integer-coded categorical variables in the PUMS data, and other common operations on this dataset.

Using `PUMSutils` the above example looks like this:

```{r estimate}
estimate(wa.house16)
```

In this case, there is no real difference. 
Computing a standard error using the weight replicates is more involved, 
and requires using the 80 replicate weights.
See the PUMS technical documentation for details.

Using `PUMSutils` the standard errors are computed with a function `acs.se`.

```{r se}
acs.se(estimate, wa.house16)
```

The value given above in `wa.gold16` for `pums_se_16` was 675.

The ACS PUMS data contains lots of integer-coded categorical values.
The PUMSUtils package contains functions for translating those to a readable form. 
It also contains functions for computing statistics over groups of data.
Here we compute totals for owned, rented and vacant housing in Washington State, using the 'TEN' field. Compare these results to the selection from `wa.gold16` above.

```{r group}
wa.house16$Own.Rent <- acs.translate(wa.house16, 
                                   'TEN', 
                                   c(1:4, NA), 
                                   c('own', 'own', 'rent', 'rent', 'vacant'))   
group.count(wa.house16, 'Tenure', 'Own.Rent')
```

The values of the 'TEN' field are:

1. owned with mortgage
2. owned free and clear
3. rented for cash
4. occupied w/o payment
 
In general, you will have to look at the data dictionary for the definitions.
It is on the Census Bureau's [technical documentation page](http://www.census.gov/programs-surveys/acs/technical-documentation/pums/documentation.html).
For the `wa.house16` data, you can use `?wa.house16` to get help for this data set.

Check the `PUMSutils` documentation for the remaining functions.
